---
title: "KNB Notebook"
output: html_notebook
---

# KNB

> The Knowledge Network for Biocomplexity (KNB) is an international repository intended to facilitate ecological and environmental research.

You don't have to become a user to access data in KNB.

# Accessing KNB Data
You could either search and download that specific data or use their [API](https://knb.ecoinformatics.org/api).

To use their API, you need to install the [rDataone](https://github.com/DataONEorg/rdataone) package in Java, Python or R. I chose R, using the following:
```{r, eval = FALSE}
install.packages("dataone")
library(dataone)
```
However, when installing the package, I came across this error:

> Configuration failed because redland was not found.

Following its error message, I installed "redland" using the command (MacOS):
```{r, eval = FALSE}
$ brew install redland 
```
And then I'm able to install the dataone package.



```{r, eval = FALSE}
cn <- CNode("PROD")
mn <- getMNode(cn, "urn:node:KNB")
mySearchTerms <- list(q="keywords:climate",
                      fl="id,title,dateUploaded,abstract,size",
                      sort="dateUploaded+desc")
result <- query(mn, solrQuery=mySearchTerms, as="data.frame")
result
result[1, c("id", "title")]
id <- result[1,'id']



metadata <- rawToChar(getObject(mn, id))
doc = xmlRoot(xmlTreeParse(metadata, asText=TRUE, trim = TRUE, ignoreBlanks = TRUE))
tf <- tempfile()
saveXML(doc, tf)
# file.show(tf)
df
```
```{r, eval = FALSE}
dataRaw <- getObject(mn, "knb.92077.1")
dataChar <- rawToChar(dataRaw)
theData <- textConnection(dataChar)
df <- read.csv(theData, stringsAsFactors=FALSE)
df[1,]
```
library(XML)
library(methods)
df <- xmlToDataFrame("tf")

# Questions
- What is XML? How can I make a data frame from it?

# My Exploration
Since there are a lot of XML files in the databases, I need to learn about XML, which looks really daunting but after a bit reading of [w3schools](https://www.w3schools.com/xml/xml_whatis.asp), I found it fairly manageble. 

Well after reading w3schools, things get a lot more complicated. It tells me to write a Java class with all the readers, scanners and stuff. I think I'll save this after I finish my programming class in Java this semester.

So I googled "import data from XML using R", yes, [lifesaver](https://www.youtube.com/watch?v=1cM_ZNZ9hhE), which is a Youtube vedio, comes up. R package makes it so much easier. Hence I may consider also the Google Earth Engine to be included in my exploration because I want to learn more about data processing. Or maybe not.

Installing `XML`
```{r, eval = FALSE}
install.packages("XML")
install.packages("methods")
```



