---
title: "KNB Notebook"
output: html_notebook
---

# KNB

> The Knowledge Network for Biocomplexity (KNB) is an international repository intended to facilitate ecological and environmental research.

You don't have to become a user to access data in KNB.

# Accessing KNB Data
You could either search and download that specific data or use their [API](https://knb.ecoinformatics.org/api).

To use their API, you need to install the [rDataone](https://github.com/DataONEorg/rdataone) package in Java, Python or R. I chose R, using the following:
```{r, eval = FALSE}
install.packages("dataone")
```
However, when installing the package, I came across this error:

> Configuration failed because redland was not found.

Following its error message, I installed "redland" using the command (MacOS):
```{r, eval = FALSE}
$ brew install redland 
```
And then I'm able to install the dataone package.


To search and get the result, follow the example on the [website](https://github.com/DataONEorg/rdataone) or use `??dataone` to see manual. 
```{r, eval = FALSE}
# load the package first
library(dataone)

# search the KNB database with the keyword "climate"
cn <- CNode("PROD")
mn <- getMNode(cn, "urn:node:KNB")
mySearchTerms <- list(q="abstract:climate",
                     rows= "500", fl="id,title,dateUploaded,abstract,size",
                      sort="dateUploaded+desc")

# show the search result as a data frame
result <- query(mn, solrQuery=mySearchTerms, as="data.frame")
head(result, 10)

# get the id of the first result
result[1, c("id", "title")]
id <- result[1,'id']
```

Now get the data from the database using `XML` and `dataone` packages.
```{r, eval = FALSE}
library(XML)
metadata <- rawToChar(getObject(mn, id))
doc = xmlRoot(xmlTreeParse(metadata, asText=TRUE, trim = TRUE, ignoreBlanks = TRUE))
tf <- tempfile()
saveXML(doc, tf)
file.show(tf)
library(methods)
xml_tf <- xmlParse(tf)
```


Write a data frame to a .csv file.
```{r, eval = FALSE}
dataRaw <- getObject(mn, "knb.92077.1")
dataChar <- rawToChar(dataRaw)
theData <- textConnection(dataChar)
df <- read.csv(theData, stringsAsFactors=FALSE)
write.csv(df, file = "~/Documents/knb-susquehanna-river-flow.csv")
```



Since there are a lot of XML files in the databases, I need to learn about XML, which looks really daunting. [w3schools](https://www.w3schools.com/xml/xml_whatis.asp) is a good website for introduction to XML, yet its method of reading xml files is pretty complicated.

So I googled "import data from XML using R", and found the the Youtube [tutorial](https://www.youtube.com/watch?v=1cM_ZNZ9hhE) of using XML package. 

Installing `XML`
```{r, eval = FALSE}
install.packages("XML")
install.packages("methods")
```

```{r, eval = FALSE}
library(XML)
library(methods)
df <- xmlToDataFrame("tf")
```
When I try to change the the xml file to a data frame it gives me an error. So I'll keep working on xml files. 

Potential solutions:
https://stackoverflow.com/questions/17198658/how-to-parse-xml-to-r-data-frame
history
python xml
